Reanalysis Step 3: Comprehensive Imputation Diagnostics

### 2025-08-30 reanalysis-3-imputation_diagnostics.qmd
diag vis

## 5. Temporal Consistency

@@

## 6. Multiple Imputation Variance

@@

## 7. Summary Assessment

>>>>

## 5. Direction of Change by Missingness Pattern

```{r change-direction-analysis}
# Create detailed groups based on what data was available
direction_df <- imputed_datasets[[1]] |>
  mutate(
    change_baseline_outcome = outcome_VA_logMAR - baseline_VA_logMAR,

    # Define groups based on original missingness
    imputation_group = case_when(
      cohort_id == "submitted" ~ "Complete (Observed)",
      cohort_id == "excluded" & !is.na(df_original$intermediate_VA_logMAR[row_number()]) ~ "Imputed (had intermediate)",
      cohort_id == "excluded" & is.na(df_original$intermediate_VA_logMAR[row_number()]) ~ "Imputed (no intermediate)",
      TRUE ~ "Other"
    ),

    # Categorize changes
    change_direction = case_when(
      change_baseline_outcome < -0.1 ~ "Improved",
      change_baseline_outcome > 0.1 ~ "Worsened",
      TRUE ~ "Stable"
    )
  ) |>
  filter(!is.na(change_baseline_outcome))

# Calculate proportions by group
direction_summary <- direction_df |>
  group_by(imputation_group, change_direction) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(imputation_group) |>
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1)
  ) |>
  ungroup()

# Create table
direction_table <- direction_summary |>
  select(imputation_group, change_direction, n, pct) |>
  pivot_wider(
    names_from = change_direction,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = 0)
  )

kable(direction_table,
      caption = "Direction of vision change by data availability group") |>
  kable_styling(full_width = FALSE)

# Visualization - stacked bar chart
p_direction <- ggplot(direction_summary, 
                     aes(x = imputation_group, y = pct, fill = factor(change_direction,
                                                                      levels = c("Improved", "Stable", "Worsened")))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = paste0(pct, "%")), 
           position = position_dodge(width = 0.9),
           vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Improved" = "#2ca02c", 
                              "Stable" = "#ff7f0e", 
                              "Worsened" = "#d62728"),
                    name = "Vision Change") +
  labs(
    title = "Direction of Vision Change by Data Availability",
    subtitle = "Comparing complete vs imputed groups",
    y = "Percentage (%)",
    x = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  ylim(0, max(direction_summary$pct) * 1.15)  # Add space for labels

print(p_direction)

# Chi-square test for independence
chi_table <- table(direction_df$imputation_group, direction_df$change_direction)
chi_test <- chisq.test(chi_table)

cat(sprintf("\nChi-square test: χ² = %.2f, p = %.4f\n", 
            chi_test$statistic, chi_test$p.value))
if(chi_test$p.value < 0.05) {
  cat("The proportions of improved/stable/worsened differ significantly across groups.\n")
} else {
  cat("The proportions are similar across groups.\n")
}
```

## 6. Temporal Consistency

@@

## 7. Multiple Imputation Variance

@@

## 8. Summary Assessment
